\documentclass[12pt]{article}

\title{Insert Report Title Here}
\author{Benjamin White-Horne \\ \emph{Oxford University}}

\newcommand{\row}[1]{\texttt{#1}}

\newcommand{\br}[0]{\vspace{10pt} \noindent}

\begin{document}

% Title page & T.o.C.
\maketitle
\pagebreak
\tableofcontents
\pagebreak

\section{What is Composing?}

\section{What is my Project Trying to Do?}

The purpose of the project is to build an application which \emph{helps} composers create better
change ringing compositions with as little change to their workflow as possible.  This aim is
roughly analogous to what applications like MuseScore and Sibelius do for musical composers --- the
goal is to create some form of `helpful paper' which allows the composer to experiment more freely
by providing instant feedback on everything that the composer wants to know about their composition.

Within this overall goal, here are some more specific things that I would expect from such an
application (most of these come from observations of existing programs, conversations with other
composers, and knowledge gained from experimenting with prototypes early on in this project):

\begin{enumerate}
    \item \textbf{Ease of Use}: The application should be made so that anyone can install and use
        it.
    \item \textbf{Completeness}: Everything considered ``change ringing'' should be representable in
        the application.  Completeness is defined by the Framework for Change Ringing.
    \item \textbf{Generality}: If a composer has an idea for a composition then this should be
        representable in the application, regardless of how complete the idea is.
    \item \textbf{Incremental}: The application should allow the user to incrementally build up
        their composition, in whatever order they want, starting wherever they want to.
    \item \textbf{Instant}: Feedback on important measures like truth, music content and length
        should always update instantly whenever the composer changes their composition.
    \item \textbf{Visual}: The application should display the composition to the user in way that
        makes it easy for the user to visually understand the composition they are working on.
    \item \textbf{View-Independent}: The application should allow the user to choose how they view
        the composition they are working on (and this should be configurable whilst midway through a
        composition).
\end{enumerate}

\section{Why Would This be Useful?}

There are several reasons why mechanical assistance would be useful for change ringing composers in
particular.

Firstly, change ringing compositions are at their core simply sequences of permuatations, and are
therefore extremely ameanable to mathematical and computerised analysis.  Also, nearly all
compositions are fundamentally designed for human consumption, putting very low upper bounds on the
potential size of the compositions which computers have to deal with (there aren't any towers with
more than 16 (ringable) bells and humans can ring about 1,500 rows per hour, so perhaps 99.99\% of
compositions have at most 16 bells and are shorter than 10,000 rows --- and at least 99\% of these
have at most 12 bells and not much over 5,000 changes).

Secondly, mathematical properties of change ringing compositions (specifically the `truth' of a
composition) are very, very important.  For the purposes of this report, a composition is `true' if
every row contained in the composition is unique, otherwise it is `false'.  Maintaining truth of a
composition is absolutely vital, since a performance of a false composition is considered invalid
and therefore is a waste of the ringers' time.  Checking proof is also extremely easy to do with a
computer (especially due to the bounds on the size of the problem) and as a result, even the
earliest computers have been used to check compositions for truth (the story goes that one of the
first programs run on the Manchester Baby was for composition proving, but this is hard to verify).
With the increasing availability of processing power computers have been used to entirely generate
compositions with minimal input from a human, but there is still something missing from computer
generated compositions.

What is currently missing, however, is a program which allows a human composer to combine the
strengths of computers and humans (computers are incredibly fast and accurate for menial grunt-work
tasks such as truth checking, whereas a skilled human composer has a good knowledge of what makes a
`good' composition and what they are trying to achieve in each specific case).  Therefore, I propose
an application which allows a human composer to get instant feedback (truth, music content, length,
etc.) about their composition \emph{whilst they are making it}.  This support for `partial'
compositions (i.e.\ compositions which are simply chunks of rows which will be rung in some
yet-to-be-defined order) is the primary motivation for building a new program rather than simply
modifying an existing one.

\section{What has been done so far?}

In this section, I will go through existing tools and review them against the goals set above.

\subsection{Composition Library}

This is a strange one, since Composition Library (shortened to `CompLib' for the rest of this
document) is not trying to be a composition editor, but simply an attempt to build a complete
central library of all the compositions that are known to ringing.  However, it does have a
composition editor in order to allow people to add new compositions to the library, and this
editor has a lot of features useful for experimenting with compositions (despite being designed for
a different purpose).  In fact, it does well enough that I think it's probably the best composing
tool out there right now, and I have made several compositions using only CompLib, paper and a
pencil for assistance.  More than anything, I think the fact that the best tool right now isn't even
designed for the purpose highlights how much promise a custom-designed composing tool holds.

% TODO: Add numbers to these

\paragraph{Ease of use:}  CompLib does really well for this.  It is really easy (in fact trivial) to
install --- it is simply a website, and therefore has no software dependencies except a browser
(which will be how most people download software anyway).  It does need an account in order to make
compositions, but this is not really an issue.

\paragraph{Incremental:}  CompLib can't handle partial compositions at all; compositions must start
(and preferrably end) in rounds.  If the composition does not come round (as is the case for almost
all unfinished compositions), then CompLib generates thousands of repeated rows on the end of your
composition in the hope that the comp will come round, to the detriment of performance and
usability.

\paragraph{Instant:}  CompLib does reasonably well.  If the composition is a round block, then it
gives a good visual indication of where the falseness occurs.  However, it isn't perfect.  The main
problem is that all the row generation and proving is handled server-side, which means that any
change to the composition has to travel across the internet before the user gets any feedback.
Also, there is a related concurrency bug where making too many changes in quick succession will
cause some of them to be overwritten and lost.

\paragraph{Visual \& View-Independence:}  CompLib again does reasonably well (and it is flawed only
because it is designed for inputting and later learning already finished compositions).
CompLib allows the user to choose between a range of several ways of viewing their composition (even
when it doesn't come round), and all of these are very easy to read.  On the other hand, the
blueline view doesn't hightlight musical rows, which I think would be beneficial even for CompLib's
use case of people learning comps.  Also, if you want to see more detail of one segment of a
composition, there is no way to `unfold' only part of it --- you have to switch to an entirely new
view and then find that location again.

\begin{enumerate}
    \item If the composition is complete, CompLib has a really nice music breakdown available to see
        if your goals for the composition has been met.  However, this is only available for
        complete compositions and viewing it requires saving and reopening the composition which is
        not ideal.
    \item CompLib's input format is very declarative --- in essence, the user enters the composition
        as a sequence of instructions for how to build the rows and these instructions are read
        in order and used to generate (`prick') the rows of your composition.  These rows are then
        used to calculate the falseness of your composition, and what will be displayed to you.
        This, however, is frustrating when you're experimenting with compositions, since a change
        near the top of a composition will cause different rows to be generated for the rest of the
        composition.  This behaviour makes perfect sense because CompLib's editor is designed for
        inputting a composition which you already know (not for experimentally building up a new
        composition) but is nonetheless makes experimentation frustrating.
\end{enumerate}

\subsection{Inpact}

Inpact is a program written by Alexander Holroyd to support the workflow of experimenting with
compositions.  It provides instant feedback on truth and music, but is visually quite clunky and
doesn't support `partial' compositions.

\paragraph{Ease of use:}

\paragraph{Incremental:}

\paragraph{Instant:}

\paragraph{Visual \& View-Independence:}

\subsubsection{Things that Inpact does well}

\begin{enumerate}
    \item It has a folding interface --- the user can fold away sections of the composition in order
        to get a higher-level overview of the composition as a whole.
    \item It doesn't require that your composition is complete before giving feedback about truth or
        music content.
    \item It has customisable music scoring --- the user decides what music they care about.
\end{enumerate}

\subsubsection{Things that Inpact doesn't do well}

\begin{enumerate}
    \item Doesn't handle `partial' compositions --- compositions have to be a single block starting
        at rounds, but don't have to come round.
    \item There isn't a visual way to see the music in the composition.
\end{enumerate}

\section{Initial Design}

Now that we've taken a look at the state of the art, it's time to start making design decisions
about this project.  To do this, I will lay out my initial design and why I think it will satisfy
each of the design goals.

\subsection{The Application Model}

This is the high-level model that should define how the application should behave (even if the
implementation details are different).  This model contributes to most of the learning curve for new
users and therefore should be as simple as possible without losing expressivity.  I'm going to
describe the model in terms of Haskell datatypes, using names identical to those used in the
project's source code (the actual implementation will look largely similar due to Rust being heavily
inspired by Haskell).

\subsubsection{Specification vs Derived State}

In order for the model to be as simple as possible, it is important to differentiate the
\emph{specification} from any \emph{derived state} generated from it.  As a toy example, let us
consider two integers ($a$ and $b$) and their sum and product ($a + b$ and $ab$) as the total
information required to represent a particular state.  This overall state consists of 4 pieces of
information ($\{a, b, a + b, ab\}$) but it's easy to see that the inclusion of $a + b$ and $ab$ does
not add any extra information.  Therefore, we can say that $\{a, b\}$ is the \emph{specification} of
$\{a, b, a + b, ab\}$, whereas $\{a + b, ab\}$ is the \emph{derived state}.  This dependence also
means that we don't have to store any derived state in the undo history, drastically reducing the
memory footprint of storing long undo histories (provided that the derivation process is not
prohibitively expensive --- though in the case of this project it is quite cheap in both time and
memory).  Plus, the mental load on the user is reduced because they only have to reason about the
specification and the behaviour of the derived state should be intuitive given the changes to the
specification.

For this project, we have a few requirements which will constrain the model used (namely the
completeness and generality goals).  Firstly, to satisfy generality the application must be able to
express composers' ideas about 

\subsection{Ease of use}

The best installation process is no installation process, so my application will run as a single
static web page with no external dependencies or accounts required.  Having no dependence on a
server both obviates the need for the user to make an account and also makes it trivial to deploy
using GitHub pages or a similar service.  Also from a user-experience (UX) perspective, HTML canvas
rendering (while extremely inefficient) is fast enough to feel instant and consistently looks good
across all platforms and browsers.  Also, much efficiency can be gained by rendering sections of the
composition into caches, which would be a relatively straightforward optimisation.

However, writing an application as a static web page is not without its challenges.

\br{}Firstly, JavaScript (though good for user interaction code) is both difficult to maintain
at any kind of scale/complexity and also border-line impossible to get consistently good perforance
when doing bulk number crunching.  Both of these are very important to me; composition generation
requires quite a lot of fairly complex code that has to do a huge amount of number crunching in a
short space of time.  My solution to this is to write the code with a client/server architecture,
where there is a thin client written in JavaScript (to handle user interaction and rendering) and
the server is written in Rust and compiled to WebAssembly.  This way, as much of the business logic
is kept in Rust (which is extremely fast and easy to maintain) but there is no network delay between
the user's inputs and visual feedback.

This approach has been very promising performance-wise: as of the first prototype, the latency
between a user changing the composition and the result appearing on screen is 3ms for a composition
with 2000 rows of 8 bells each.  This processess involves no caching --- the entire composition is
completely expanded, proved, music checked, serialised to JSON, copied to JavaScript and
deserialised in the 3ms.  This could definitely be optimised, but we are so far under the 100ms
limit of feeling instant that this would be a waste of my time.

\br{}The second, more problematic limitation of working inside a browser is that of storage.
Code running in a browser cannot (for security reasons) access the file system unless the user
specifically uploads or saves a file, and therefore accessing the file system automatically when the
user types `ctrl-S' is not an option.  Also, the best thing for the user's experience is to be able
to close and reopen the page and have the page load into exactly the state it was in before closing
(complete with as much undo history as possible).

Fortunately for this project, there is some storage that web pages can use to store persistent data
without the user's input, and those are cookies.  Unfortunately, cookies are severly size limited
--- a web page can only expect to be able to save about 30 cookies per page, each of which can be at
most 4096 bytes, so we have 122,800 bytes in which to save all the persistent state we need.  In
reality, however, it's not a good idea to run too close to this limit because the page may have to
store other cookies and these shouldn't suddenly fail to save.  Saving just the current undo
snapshot is fine, but I want to save the undo history which will require a complicated
serialisation/deserialisation algorithm.  I have chosen not to implement this for the prototype,
because I suspect that I will need to change the internal representation a lot and doing so will
break any serialised data.

\subsection{Incremental}

This is where the proper design starts.  In order to allow the application to be intuitive, there
must be a relatively simple underlying model of how the user should think about the composition they
are editing.

A composition is an unordered counted set (i.e.\ a ``bag'') of ``fragments''.

\end{document}
